---
title: "Coding Style"
output:   
    ioslides_presentation:
      highlight: pygments
      smaller: yes
      theme: readable
      widescreen: no
      css: mainPresentation.css
---


## Why is the following code problematic?


```{r, eval=FALSE}
T<-matrix(rnorm(10000), 100, 100)
data = matrix(NA,100,100)


for(i in 1:100){
for (j in 1:100){
  data2[i,j] <-sum(T[,i]*T[,j])
}
}

foo = function(c,t){
if(c == t) {y = rbinom(1,1,.5)}
else {y = rbinom(1,1,.6)}
y  
}
```

## Issues

- Vague object names
- Inappropriate object names
- Inconsistent assignment
- Unclear arguments
- Unnecessary looping
- Problematic spacing
- No comments
- Magic numbers


## Comparison

```{r, eval=FALSE}
### The following generates the crossproduct of a matrix

# sample size
N = 100

# simulate data from draws from a normal distribution
set.seed(123)  # if desire reproducibility
randNormMat = matrix(rnorm(N*N, ncol=N, nrow=N))

crossprodMat = crossprod(randNormMat)

### example function serving no purpose

# takes two integer vectors, draws a random sample from both. If the two values
# are equal draws from binomial size=1 with first probability, else uses the
# second probability

exampleFunction = function(integerVector1, integerVector2, prob1, prob2){
  sampleInt1 = sample(integerVector1, 1)
  sampleInt2 = sample(integerVector2, 1)
  
  if(sampleInt1 == sampleInt2) {
    randomBinomDraw = rbinom(1, size=1, prob=prob1)
    } else {
      randomBinomDraw = rbinom(1, size=1, prob=prob2)
      }
  randomBinomDraw
}
```


## Good coding practices
- Consistent coding style, including spacing, assignment etc.

- Comment early and often
    - Focus on why not what

- Strive for clarity
    - Assume you'll need the code again a year from when you are writing it- will you know what you've done?
    
