---
title: "Dealing with Data"
author: "MC"
output:   
    ioslides_presentation:
      highlight: zenburn
      smaller: yes
      theme: readable
      widescreen: no
      css: mainPresentation.css
---
```{r, echo=FALSE}
library(knitr)
opts_knit$set(width=120, message=FALSE)
```
## Outline

- Two approaches
    - Base R
    - Piping
- Filtering
- Selecting
- Slicing
- Using non-standard data objects


# Indexing
## Base R Indexing
Base R has a very flexible/powerful indexing system with which to get at your data.

Consider the following simple example, where we want values of x greater than 1

```{r indexExample, echo=-1, cache=TRUE}
set.seed(1234)
x = rnorm(10)
x[x>0 & x<1]
```

A more complicated approach. Values of y in which the absolute residuals of a regression on x are greater than 1.

```{r indexExample2, eval=FALSE}
y[abs(resid(lm(y~x))) > 1]
```

This is not the way you should be doing things, but serves as an example of the possibilities

## Base R Indexing

Indexing

Lists (or dataframes)

- ['name'] [number] to extract list elements (including data.frames)
    - myList['myelement']
    - [[ can select a single element, where [] can return multiple
- \$name also
    - myData$myvariable
    
Matrices (or dataframes)

- [rows, columns] or ['rownames', 'columnnames']
    - myData[1:3, 6:10]
    - myData[,'var1']
    
see ?Extract


## Base R Indexing

For this exercise, you will work with one of the many datasets that come with a basic R installation (?datasets), but note that packages typically come with their own as well.

For the following:
Examine the **iris** dataset using the str (structure) function.  You do not have to do anything to get the iris data, it's an object in the base R environment.

Create a separate object that just contains the the Species column of data

Create an object containing the 10-15th rows

Note the Petal Width of the 100th observation via indexing

Bonus: subset the data to only the virginica species.  Hint- use ==


# Piping

## Piping
Piping is an alternative way to successively perform data manipulation and indexing (mainly through magrittr package).

- The most common way to do so is with this symbol %>%
- However, there are other ways to do so, and some with other packages


In conjunction with dplyr, it can make data manipulation and wrangling straightforward and with very clear code.


## Piping
Install the rvest package (read html functionality).

```{r lePipe, cache=TRUE}
# ☻ ☻ ☻
library(dplyr); library(rvest); library(stringr)
link ="http://www.inflationdata.com/inflation/Inflation_Rate/HistoricalInflation.aspx"
html(link) %>%                                        # read a webpage
  html_table(header = TRUE) %>%                       # grab all the tables
  `[[`(1) %>%                                         # grab the first table
  select(Year, Ave.) %>%                              # select specific columns
  rename(Ave = Ave.) %>%                              # rename
  filter(Year >= 2010) %>%                            # select certain rows
  mutate(AveNum = str_extract(Ave, '[0-9].[0-9]+'),   
         AveNum2 = as.numeric(str_replace(Ave, ' %', '')))  # create new variables
```

## Piping

Note that you can assign the results of the piped operations to create an object

```{r pipe2object, eval=FALSE, cache=TRUE}
myObjectSubset = mydata %>% 
  filter(Year > 2000)
```

Piping is especially useful for visualization, and ggvis and newer more web-directed visualization packages will typically work with standard pipe operators.

```{r ggvisPipe, fig.align='center', fig.height=2.5, fig.width=4}
library(ggvis)
iris %>% 
  ggvis(x=~Petal.Length, y=~Petal.Width) %>% 
  layer_points(fill:='#ff5500', fillOpacity:=.35, size:=25) %>% 
  layer_smooths(stroke:='darkred')
```


# Filtering

## Filtering
Filtering (or Slicing) refers to subsetting by rows.

You have already seen how to do this with base R functionality with numbered indexing

the dplyr package has two primary functions for this

 - filter 
 - slice 
 
It also has other functions that can make the code clearer and easier to work wtih

```{r filterSlice, echo=-1, cache=FALSE}
library(dplyr)
state.x77 %>% 
  as.data.frame %>% 
  filter(Frost >150) %>% 
  slice(1:3)
```

# Selecting

## Selecting

Selecting refers to the selection of columns and with dplyr again you get enhanced functionality and clarity

```{r select, cache=TRUE}
iris %>% 
  select(starts_with('Petal')) %>% 
  slice(1:2)
```

## Aggregation
Oftentimes we engage in such operations in order to obtain summary information about groups.

```{r}
iris %>% 
  group_by(Species) %>% 
  summarise_each('mean')
state.x77 %>% 
  data.frame %>% 
  mutate(region = state.region) %>% 
  group_by(region) %>% 
  tally
```


# Using non-standard data
Examples of non-standard data:

- text
- images
- spatial data (e.g. maps)
- web-oriented (json, xml)

Most of the data manipulation of such objects will be to put into a more analyzable format.

You will likely have to use package specific functions (e.g. jsonlite)

#  Playtime
Link