---
title: "Control Flow"
author: "MC"
output:   
    ioslides_presentation:
      highlight: zenburn
      smaller: yes
      theme: readable
      widescreen: no
      css: mainPresentation.css
---

## Outline

- Standard looping
- Using functionals
- Vectorized approaches
- Parallelizing




# Iterative Programming
## Standard looping
Data manipulation involves a lot of repeated operations, and R shines in this area in that it makes a lot of this easier.

You'll often see iterative programming within a function, though it can be and often is used in the normal course of programming.

The typical approach seen in other languages can be used, and sometimes is still the way to go.

Calculate column means:

```{r loop, cache=T}
# million column matrix
d = matrix(rnorm(10000000), nrow=10)    # might take a second or two
dMeans = numeric(ncol(d))               # not necessary but does speed up explicit loops to predefine the object

# The following may take up to ~ 10 seconds
for (i in 1:ncol(d)){
  dMeans[i] = mean(d[,i])
}
```

## Standard looping
One may loop:

- directly over elements (less optimal)
- over the numeric indices (as the in that demo, possibly the most common in the R community)
- over the names

## Standard looping
A more general way to loop is with a <mark class='empCode'>while</mark> in place of a for.

The loop will run until some condition is met.

Any for loop can be written as a while statement.

```{r, eval=FALSE}
n = 1
while(n <= ncol(d)){
  dMeans[n] = mean(d[,n])
  n = n + 1
}
```

## If Else
Within functions you will often see something like the following:

```{r, eval=FALSE}
if(conditionMet){
  doThis
} else if(otherConditionMet) {
  doThat
} else {
  doSomethingElse
}
```

For more on for, while, and if...else constructs see ?Control.


## Looping Alternatives: Vectorized Code
While standard looping works, other approaches are available, and some *vectorized* operations can reduce code and are often faster. 

Vectorizing your code means working on whole objects rather than iterating over individual elements.


## Looping Alternatives: Vectorized Code

Consider the following loop (type it or something similar yourself):

```{r, eval=FALSE}
a = rnorm(10)
b = rnorm(10)

for(i in seq(10)){
  message(a[i] + b[i])
}
```

The following is easier to write, clearer to read, and faster:

```{r, eval=FALSE}
a + b
```

## Looping Alternatives: Vectorized Code

There are many vectorized functions that come with base R, or are otherwise optimized such that you'll want to quickly learn them and avoid using explicit loops.

  - rowSums, colSums
  - rowMeans, colMeans
  - matrix operations (e.g. crossprod)
  - scale

## Looping Alternatives: Vectorized Code
The vectorized approach works in a lot of other places also, and is especially useful in indexing.

```{r vectorize, cache=T, results='hold'}
myvar = 1:100
myvar[sample(myvar, 50)] = NA       # insert 50 missing values
sum(is.na(myvar))                   # check how many
myvar[is.na(myvar)] = 0             # change missing to 0
any(is.na(myvar))                   # any left

1:10 >= 5                           # which of the sequence 1:10 is greater or equal to 5

rowSums(d > 0)                      # count values greater than zero for each row
```


## Looping Alternatives: Functionals
The quicker you learn the <mark class='empCode'>apply</mark> family of functions, as well as related versions in the aforementioned plyr package, the quicker your data processing will be.

Apply the <mark class='empCode'>mean</mark> function to the columns.  Similar time as a loop in this case but cleaner code. 

```{r apply, cache=T}
dmeans = apply(d, 2, mean) # the 2 specifies columns
```

Note that we supply a function as an argument to  function.

## Looping Alternatives: Functionals

As mentioned earlier, one will often use anonymous functions with apply. 

We create a simple function on the fly, without assignment.

```{r anonymous, cache=TRUE}
apply(d, 1, function(x) any(x>5))
```

## Looping Alternatives: Functionals

The apply is a family of functions:

- <mark class='empCode'>lapply</mark> (will return list)
- <mark class='empCode'>sapply</mark> and vapply for vectors and lists (can return list or vector/matrix)
- <mark class='empCode'>tapply</mark> for groupwise applications
- <mark class='empCode'>mapply</mark> for multi-argument passing (for example to do something with 2 lists of input)
- <mark class='empCode'>replicate</mark> to do something N times

They also have parallelized versions, which means even if they weren't any faster than the standad loop they could be greatly sped up beyond the base speed.

## Looping Alternatives: Functionals
plyr, dplyr, and tidyr packages have functions with similar capabilities but:

  - are often faster
  - allow for easier management of input to output
  - often 'just work'
  
```{r ddply, eval=FALSE}
library(plyr)
aaply(d, 1, mean)                  # array to array
adply(d, 1, mean,  .parallel = T)  # array to dataframe in parallel
```



# Your turn
## Your turn

Using either of the two matrices below (or create one yourself), perform a row or column operation with two different approaches- one using an explicit loop, one of which uses apply-like functionality.

Examples: if any in a row equal some value, any columns with values greater than some other value.

Suggestions: 

  - Calculations like <mark class='empCode'>rowMeans</mark>, <mark class='empCode'>colSums</mark>
  - & | ! (and and or operator, ?Logic)
  - Logical Operators <mark class='empCode'>== != > < <= >= </mark> (?Logic ?Comparison for more info)
  - <mark class='empCode'>any</mark> <mark class='empCode'>which</mark> <mark class='empCode'>all</mark>
  
Hint: first decide on your problem, keep as simple as need be. Then write pseudo-code for the anonymous function part if using one.  Then try to convert to a working example.

If you want to try something that isn't clear from the above suggestions just ask.

Example:


```{r}
nums = matrix(sample(1:3, 9, replace = T), 3, 3)
lets = matrix(sample(letters, 9, replace = T), 3, 3)
```

