---
title: "Modeling Basics"
author: "MC"
output:   
    ioslides_presentation:
      highlight: zenburn
      smaller: yes
      theme: readable
      widescreen: no
      css: mainPresentation.css

---

## Modeling

After all that we've gone through at this point, modeling will be the easy part for you.  

The biggest hurdle in modeling is preparing the data, and the better you are at programming:

- the faster it will go
- the more you can explore
- the more you can spend time learning other things (like new modeling packages)

Once complete, you now have thousands of packages to choose from for modeling.


## Modeling

The standard format for regression modeling can serve as baseline code for modeling in other packages.

Most take a formula and data argument. 
- a regression of y on x and z

```{r lmExample, eval=FALSE}
lm(y ~ x + z, data=myData)
```

Some might take matrix/vector arguments.

```{r lmfitExample, eval=FALSE}
lm.fit(X, y)          # lm.fit is the workhorse of lm
```


Unsupervised methods usually will need a matrix argument 

- sometimes a dataframe is ok

```{r pca, eval=FALSE}
princomp(X)
```


## Modeling

And then there are variations on these themes, but those constitute the vast majority of standard and complex models.  

In addition, they all come with additional arguments that you'll need to familiarize yourself with before using the function

```{r lmAlt, eval=FALSE}
glm(art ~ ., data=bioChemists, family = 'poisson')   # number of articles predicted by everything else

library(pscl)
hurdle(art ~ ., data=bioChemists, dist = 'poisson')  # hurdle model to deal with excess zeros
```

```{r lmerExample, eval=FALSE}
library(lme4)
lmer(Reaction ~ Days + (1|Subject), data=sleepstudy)  # a mixed model with a random intercept for Subject
```



