---
title: "Simple programming"
author: "MC"
output:   
    ioslides_presentation:
      highlight: zenburn
      theme: cosmo
---

## Outline
  - Creating and manipulating objects
    
    - Classes of objects
  - Data Structures
  - Using functions
  
  
  
## Creating and manipulating objects

In R, you will constantly create *objects* and manipulate objects with *functions*.

Objects can be anything, and this will become clear as we continue through the course.

We will use many functions over the course of the day, including building our own.



## Data Structures

The most common data structures used can be seen to posses

One type only

  - Vector (may be of length 1)
    - numeric, integer, character and logical are the most commonly used
    - In addition, a special class of the integer type are factors
    
  - Matrix
    - collection of vectors
    
  - Array
    - beyond 2 dimensions

## Data Structures
Multiple types

  - Lists
    - Dataframes
      - Perhaps most commonly used class of lists
      
Dataframes are what you've been using if you have experience with other statistical packages or programming languages.

Typical use is that each column represents attributes, targets etc., while rows represent observations (e.g. people, time points)



## Using functions
R is a functional language.

Functions are objects that take some input and return a **value**

They are highly flexible, in that the nature of the input can range from nothing to a couple dozen ***arguments***, values can be lists of modeling output, a plot, or even another function.


## Using functions


```{r funcDemo, eval=F}
myfunction(x=someX, y=someY)
```


## Using functions

Use the matrix function to create a matrix, use c to concatentate values, use : to create sequences, and finally provide arguments pertaining to the number of rows and columns.

```{r createMatrix}
myMatrix = matrix(c(1:3, 4:6, 7:9), nrow=3, ncol=3, byrow=T)
```

Try creating a dataframe version of the myMatrix object, called myDF.  To convert a matrix object to a dataframe, use the **as.data.frame** function on it.


## Using functions

While most functions are unique, many are generic and have ***methods*** that will allow the function to operate differently for different classes of objects. The following demonstrates this with the commonly used **summary** function.

```{r methods}
methods(summary)
```



